# Limits {#sec:limits}



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/WUvTyaaNkzM")
```



*Calculus* means "a method of calculation or reasoning." When one
computes the sales tax on a purchase, one employs a simple calculus.
When one finds the area of a polygonal shape by breaking it up into a
set of triangles, one is using another calculus. Proving a theorem in
geometry employs yet another calculus.

Despite the wonderful advances in mathematics that had taken place into
the first half of the $17^\text{th}$ century, mathematicians and
scientists were keenly aware of what they *could not do.* (This is true
even today.) In particular, two important concepts eluded mastery by the
great thinkers of that time: area and rates of change.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/HQ6SqbaFLvo")
```



Area seems innocuous enough; areas of circles, rectangles,
parallelograms, etc., are standard topics of study for students today
just as they were then. However, the areas of *arbitrary* shapes could
not be computed, even if the boundary of the shape could be described
exactly.

Rates of change were also important. When an object moves at a constant
rate of change, then "distance = rate $\times$ time." But what if the
rate is not constant – can distance still be computed? Or, if distance
is known, can we discover the rate of change?




```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/I7guga5Ych8")
```


It turns out that these two concepts were related. Two mathematicians,
[Sir Isaac Newton](https://plato.stanford.edu/entries/newton/) and [Gottfried Leibniz](https://plato.stanford.edu/entries/leibniz/), are credited with independently
formulating a system of computing that solved the above problems and
showed how they were connected. Their system of reasoning was "a"
calculus. However, as the power and importance of their discovery took
hold, it became known to many as "the" calculus. Today, we generally
shorten this to discuss "calculus."

The foundation of "the calculus" is the *limit.* It is a tool to
describe a particular behavior of a function. This chapter begins our
study of the limit by approximating its value graphically and
numerically. After a formal definition of the limit, properties are
established that make "finding limits" tractable. Once the limit is
understood, then the problems of area and rates of change can be
approached.

```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/37n0cZn6Lyc")
```



## An Introduction To Limits {#sec:limitintro}


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/qwm3wAoMn6w")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/rG3XVvFIZPY")
```

We begin our study of *limits* by considering examples that demonstrate
key concepts that will be explained as we progress.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/__qzaSg4y1I")
```

Consider the function $y = \frac{\sin x}{x}$. When $x$ is near the value
1, what value (if any) is $y$ near? While our question is not precisely
formed (what constitutes "near the value 1"?), the answer does not seem
difficult to find. One might think first to look at a graph of this
function to approximate the appropriate $y$ values. Consider Figure
\@ref(fig:sinxoverx), where $y = \frac{\sin x}{x}$ is graphed.
For values of $x$ near 1, it seems that $y$ takes on values near $0.85$.
In fact, when $x=1$, then $y=\frac{\sin 1}{1} \approx 0.84$, so it makes
sense that when $x$ is "near" 1, $y$ will be "near" $0.84$.

```{r sinxoverx, fig.cap="Graph of $f(x)=\\sin x / x$.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/pbwtkz3r?embed")
```

Consider this again at a different value for $x$. When $x$ is near 0,
what value (if any) is $y$ near? By considering Figure
\@ref(fig:sinxoverx), one can see that it seems that $y$ takes on
values near $1$. But what happens when $x=0$? We have
$$y \rightarrow \frac{\sin 0}{0} \rightarrow \frac{0}{0} \rightarrow \text{undefined.}$$
The expression "$0/0$" has no value; it is *indeterminate.* Such an
expression gives no information about what is going on with the function
nearby. We cannot find out how $y$ behaves near $x=0$ for this function
simply by letting $x=0$.

*Finding a limit* entails understanding how a function behaves near a
particular value of $x$. Before continuing, it will be useful to
establish some notation. Let $y=f(x)$; that is, let $y$ be a function of
$x$ for some function $f$. The expression "the limit of $y$ as $x$
approaches 1" describes a number, often referred to as $L$, that $y$
nears as $x$ nears 1. We write all this as
$$\lim_{x\to 1} y = \lim_{x\to 1} f(x) = L.$$ This is not a complete
definition (that will come in the next section); this is a
pseudo-definition that will allow us to explore the idea of a limit.

Above, where $f(x) = \sin(x)/x$, we approximated
$$\lim_{x\to 1} \frac{\sin x}{x} \approx 0.84 \quad \text{ and } \quad \lim_{x\to 0}\frac{\sin x}{x} \approx 1.$$
(We *approximated* these limits, hence used the "$\approx$" symbol,
since we are working with the pseudo-definition of a limit, not the
actual definition.)

Once we have the true definition of a limit, we will find limits
*analytically*; that is, exactly using a variety of mathematical tools.
For now, we will *approximate* limits both graphically and numerically.
Graphing a function can provide a good approximation, though often not
very precise. Numerical methods can provide a more accurate
approximation. We have already approximated limits graphically, so we
now turn our attention to numerical approximations.

Consider again $\lim_{x\to 1}\sin (x)/x$. To approximate this limit
numerically, we can create a table of $x$ and $f(x)$ values where $x$ is
"near" 1. This is done in Table \@ref(tab:sinxnear1).

```{r sinxnear1, class.output="centered"}
tibble::tibble(x=c(0.9,0.99,0.999, 1, 1.001,1.01,1.1), "sin(x)/x"=sin(x)/x) %>% 
  knitr::kable(caption="Values of $\\sin(x)/x$ with $x$ near 1.",
               table.attr='class="centered"', format="html")
```


Notice that for values of $x$ near $1$, we have $\sin (x)/x$ near
$0.841$. The $x=1$ row is in bold to highlight the fact that when
considering limits, we are *not* concerned with the value of the
function at that particular $x$ value; we are only concerned with the
values of the function when $x$ is *near* 1.

Now approximate $\lim_{x\to 0} \sin(x)/x$ numerically. We already
approximated the value of this limit as 1 graphically in Figure
\@ref(fig:sinxoverx). Table \@ref(tab:sinxnear0) shows the
value of $\sin(x)/x$ for values of $x$ near 0. Ten places after the
decimal point are shown to highlight how close to 1 the value of
$\sin(x)/x$ gets as $x$ takes on values very near 0. We include the
$x=0$ row in bold again to stress that we are not concerned with the
value of our function at $x=0$, only on the behavior of the function
*near* 0.

```{r sinxnear0, class.output="centered"}
tibble::tibble(x=c(-0.1, -0.01, -0.001, 0, 0.001, 0.01, 0.1), "sin(x)/x"=sin(x)/x) %>% 
  knitr::kable(caption="Values of $\\sin(x)/x$ with $x$ near 0.",
               table.attr='class="centered"', format="html")
```


This numerical method gives confidence to say that 1 is a good
approximation of $\lim_{x\to 0} \sin(x)/x$; that is,
$$\lim_{x\to 0} \sin(x)/x \approx 1.$$ Later we will be able to prove
that the limit is *exactly* 1.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/eHx3LmrQZXM")
```

We now consider several examples that allow us explore different aspects
of the limit concept.\

### Example: Rational function {#sec:limit1}

To graphically approximate the limit, graph
$$y = \frac{x^2-x-6}{6x^2-19x+3}$$ on a small interval that contains 3. To
numerically approximate the limit, create a table of values where the
$x$ values are near 3. This is done in Figures \@ref(fig:limit1) and
Table \@ref(tab:limit1table), respectively.

```{r limit1, fig.cap="Graphically approximating the limit of a rational function.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/ahxgeetr?embed")
```

```{r limit1table, class.output="centered"}
tibble::tibble(x=c(2.9,2.99,2.999, 3, 3.001,3.01,3.1), y=(x^2-x-6)/(6*x^2 - 19*x +3)) %>% 
  knitr::kable(caption="Values of $y=\\frac{x^2-x-6}{6x^2-19x+3}$ with $x$ near 3.",
               table.attr='class="centered"', format="html")
```


The graph shows that when $x$ is near 3, the value of $y$ is very near
$0.3$. By considering values of $x$ near 3, we see that $y=0.294$ is a
better approximation. The graph and the table imply that
$$\lim_{x\to 3} \frac{x^2-x-6}{6x^2-19x+3} \approx 0.294.$$

This example may bring up a few questions about approximating limits
(and the nature of limits themselves).

1.  If a graph does not produce as good an approximation as a table, why
    bother with it?

2.  How many values of $x$ in a table are "enough?" In the previous
    example, could we have just used $x=3.001$ and found a fine
    approximation?

Graphs are useful since they give a visual understanding concerning the
behavior of a function. Sometimes a function may act "erratically" near
certain $x$ values which is hard to discern numerically but very plain
graphically. Since graphing utilities are very accessible, it makes
sense to make proper use of them.

Since tables and graphs are used only to *approximate* the value of a
limit, there is not a firm answer to how many data points are "enough."
Include enough so that a trend is clear, and use values (when possible)
both less than and greater than the value in question. In Example
\@ref(sec:limit1), we used both values less than and greater than 3. Had we
used just $x=3.001$, we might have been tempted to conclude that the
limit had a value of $0.3$. While this is not far off, we could do
better. Using values "on both sides of 3" helps us identify trends.\
\

```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/7RAiKoLCpgU")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/BYWMYisNllY")
```

### Identifying When Limits Do Not Exist {-}

A function may not have a limit for all values of $x$. That is, we
cannot say $\lim_{x\to c}f(x)=L$ for some numbers $L$ for all values of
$c$, for there may not be a number that $f(x)$ is approaching. There are
three common ways in which a limit may fail to exist.

1.  The function $f(x)$ may approach different values on either side of
    $c$.

2.  The function may grow without upper or lower bound as $x$ approaches
    $c$.

3.  The function may oscillate as $x$ approaches $c$ without approaching
    a specific value.

```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/DSIaDa_ABxo")
```


We’ll explore each of these in turn.

### Different values approached from left and right {#sec:nolimit1}

Explore why $\displaystyle\lim_{x\to 1} f(x)$ does not exist, where 
$$f(x) = \begin{cases} x^2-2x+3 & x\leq 1 \\ x & x>1 \end{cases}.$$

```{r nolimit1, fig.cap="Graphically approximating the limit of a piecewise function.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/qmysttuf?embed")
```

```{r nolimit1table, class.output="centered"}
tibble::tibble(x=1+c(-0.1,-0.01,-0.001, 0, .001,.01,0.1), y=ifelse(x<=1, x*x-x-x+3, x)) %>% 
  knitr::kable(caption="Values of $f(x)$ with $x$ near 1.",
               table.attr='class="centered"', format="html")
```



A graph of $f(x)$ around $x=1$ and a table are given in Figures
\@ref(fig:nolimit1) and \@ref(tab:nolimit1table), respectively. It is clear that
as $x$ approaches 1, $f(x)$ does not seem to approach a single number.
Instead, it seems as though $f(x)$ approaches two different numbers.
When considering values of $x$ less than 1 (approaching 1 from the
left), it seems that $f(x)$ is approaching 2; when considering values of
$x$ greater than 1 (approaching 1 from the right), it seems that $f(x)$
is approaching 1. Recognizing this behavior is important; we’ll study
this in greater depth later. Right now, it suffices to say that the
limit does not exist since $f(x)$ is not approaching one value as $x$
approaches 1.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/NnV1A1KTbg0")
```


### Example: Growth Without Bound {#sec:ex_no_limit2}

A graph and table of $f(x) = 1/(x-1)^2$ are given in Figures \@ref(fig:nolimit2) and \@ref(tab:nolimit2table), respectively. Both show that as
$x$ approaches 1, $f(x)$ grows larger and larger.

We can deduce this on our own, without the aid of the graph and table.
If $x$ is near 1, then $(x-1)^2$ is very small, and:
$$\frac{1}{\text{very small number}} = \text{very large number}.$$ Since
$f(x)$ is not approaching a single number, we conclude that
$$\lim_{x\to 1}\frac{1}{(x-1)^2}$$ does not exist.


```{r nolimit2, fig.cap="A function that grows without bound.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/q9kc2gtt?embed")
```

```{r nolimit2table, class.output="centered"}
tibble::tibble(x=1+c(-0.1,-0.01,-0.001, 0, .001,.01,0.1), y=1/(x-1)^2) %>% 
  knitr::kable(caption="Values of $f(x)$ with $x$ near 1.",
               table.attr='class="centered"', format="html")
```

### Example: Oscillation

Explore why $\displaystyle \lim_{x\to 0}\sin(1/x)$ does not exist.

```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/gGvvFX5QyjE")
```


```{r nolimit3, fig.cap="A function that oscillates at the limit point.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/ewgs887d?embed")
```

A graphs of $f(x) = \sin(1/x)$ is given in Figure \@ref(fig:nolimit3).
Finally, in Table \@ref(tab:nolimit3table), we see $\sin(1/x)$ evaluated
for values of $x$ near 0. As $x$ approaches 0, $f(x)$ does not appear to
approach any value.

It can be shown that in reality, as $x$ approaches 0, $\sin(1/x)$ takes
on all values between $-1$ and 1 infinitely many times! Because of this
oscillation,

$\displaystyle\lim_{x\to 0}\sin(1/x)$ does not exist.

```{r nolimit3table, class.output="centered"}
tibble::tibble(x=c(0.1,0.01,0.001, 0.0001,0.00001), "$\\sin(1/x)$"=sin(1/x)) %>% 
  mutate(x=format(x, scientific = FALSE)) %>% 
  knitr::kable(caption="Values of $f(x)$ with $x$ near 0.",
               table.attr='class="centered"', format="html")
```



### Limits of Difference Quotients {-}

We have approximated limits of functions as $x$ approached a particular
number. We will consider another important kind of limit after
explaining a few key ideas.

```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/2NJmd0Jrt4U")
```


Let $f(x)$ represent the position function, in feet, of some particle
that is moving in a straight line, where $x$ is measured in seconds.
Let’s say that when $x=1$, the particle is at position 10 ft., and when
$x=5$, the particle is at 20 ft. Another way of expressing this is to
say $$f(1)=10 \quad \text{ and } \quad f(5) = 20.$$ Since the particle
traveled 10 feet in 4 seconds, we can say the particle’s *average
velocity* was 2.5 ft/s. We write this calculation using a "quotient of
differences," or, a *difference quotient*:
$$\frac{f(5) - f(1)}{5-1} = \frac{10}4 = 2.5 \text{ft/s}.$$

This difference quotient can be thought of as the familiar "rise over
run" used to compute the slopes of lines. In fact, that is essentially
what we are doing: given two points on the graph of $f$, we are finding
the slope of the *secant line* through those two points. See Figure
\@ref(fig:diffquot).

Now consider finding the average speed on another time interval. We
again start at $x=1$, but consider the position of the particle $h$
seconds later. That is, consider the positions of the particle when
$x=1$ and when $x=1+h$. The difference quotient is now
$$\frac{f(1+h)-f(1)}{(1+h)-1} = \frac{f(1+h)-f(1)}h.$$

```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/lNI90Y321b0")
```

Let $f(x) = -1.5x^2+11.5x$; note that $f(1)=10$ and $f(5) = 20$, as in
our discussion. We can compute this difference quotient for all values
of $h$ (even negative values!) except $h=0$, for then we get "0/0," the
indeterminate form introduced earlier. For all values $h\neq 0$, the
difference quotient computes the average velocity of the particle over
an interval of time of length $h$ starting at $x=1$.

For small values of $h$, i.e., values of $h$ close to 0, we get average
velocities over very short time periods and compute secant lines over
small intervals. See Figure \@ref(fig:diffquot). This leads us
to wonder what the limit of the difference quotient is as $h$ approaches
0. That is, $$\lim_{h\to 0} \frac{f(1+h)-f(1)}{h} = \text{ ? }$$



```{r diffquot, fig.cap="The secant lines of $f(x)$ at a point $a$. Move the slider for $h$ to investigate the limit of the difference quotient.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/cgazabpr?embed")
```

As we do not yet have a true definition of a limit nor an exact
method for computing it, we settle for approximating the value. While we
could graph the difference quotient (where the $x$-axis would represent
$h$ values and the $y$-axis would represent values of the difference
quotient) we settle for making a table. See Figure
\@ref(tab:diffquottable). The table gives us reason to assume the
value of the limit is about 8.5.\

```{r diffquottable, class.output="centered"}
f <- function(x) -1.5*x^2 + 11.5 * x
tibble::tibble(h=c(-0.1,-0.01,-0.001, 0, .001,.01,0.1), "$\\frac{f(1+h)-f(1)}{h}$"=(f(1+h)-f(1))/h) %>% 
  mutate(h=format(h, scientific = FALSE)) %>% rename("$h$"=h) %>% 
  knitr::kable(caption="The difference quotient evaluated with $h$ near 0.",
               table.attr='class="centered"', format="html")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/uUyZHfTYlBQ")
```


Proper understanding of limits is key to understanding calculus. With
limits, we can accomplish seemingly impossible mathematical things, like
adding up an infinite number of numbers (and not get infinity) and
finding the slope of a line between two points, where the "two points"
are actually the same point. These are not just mathematical
curiosities; they allow us to link position, velocity and acceleration
together, connect cross-sectional areas to volume, find the work done by
a variable force, and much more.

In the next section we give the formal definition of the limit and begin
our study of finding limits analytically. In the following exercises, we
continue our introduction and approximate the value of limits.\






## Epsilon-Delta Definition of a Limit {#sec:limitdef}


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/OGvDIXuWn0g")
```


This section introduces the formal definition of a limit. Many refer to
this as "the epsilon-delta," definition, referring to the letters
$\epsilon$ and $\delta$ of the Greek alphabet.\
Before we give the actual definition, let’s consider a few informal ways
of describing a limit. Given a function $y=f(x)$ and an $x$-value, $c$,
we say that "the limit of the function $f$, as $x$ approaches $c$, is a
value $L$":

1. if "$y$ tends to $L$" as "$x$ tends to $c$."

2. if "$y$ approaches $L$" as "$x$ approaches $c$."

3. if "$y$ is near $L$" whenever "$x$ is near $c$."

The problem with these definitions is that the words "tends,"
"approach," and especially "near" are not exact. In what way does the
variable $x$ tend to, or approach, $c$? How near do $x$ and $y$ have to
be to $c$ and $L$, respectively?\
The definition we describe in this section comes from formalizing
**3**. A quick restatement gets us closer to what we want:

> If $x$ is within a certain *tolerance level* of $c$, then the
    corresponding value $y=f(x)$ is within a certain *tolerance level*
    of $L$.



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/npoSY-AFvOY")
```



The traditional notation for the $x$-tolerance is the lowercase Greek
letter delta, or $\delta$, and the $y$-tolerance is denoted by lowercase
epsilon, or $\epsilon$. One more rephrasing
nearly gets us to the actual definition:

>   If $x$ is within $\delta$ units of $c$, then the corresponding value
    of $y$ is within $\epsilon$ units of $L$.

We can write "$x$ is within $\delta$ units of $c$" mathematically as
$$|x-c| < \delta, \qquad \text{which is equivalent to }\qquad c-\delta < x < c+\delta.$$
Letting the symbol "$\longrightarrow$" represent the word "implies," we
can rewrite our statement as
$$|x - c| < \delta \longrightarrow  |y - L| < \epsilon 
\qquad \textrm{or} \qquad c - \delta < x < c + \delta \longrightarrow L - \epsilon < y < L + \epsilon.$$
The point is that $\delta$ and $\epsilon$, being tolerances, can be any
positive (but typically small) values. Finally, we have the formal
definition of the limit with the notation seen in the previous section.


> 
```{definition, name="The Limit of a Function $f$ at a point $c$.", label="limit", echo=TRUE}
Let $I$ be an open interval containing $c$, and let $f$ be a function defined on $I$, except possibly at $c$. The **limit of $f(x)$, as $x$ approaches $c$, is $L$**, denoted by  
$$\displaystyle \lim_{x\rightarrow c} f(x) = L,$$
means that given any $\epsilon > 0$, there exists $\delta > 0$ such that for all $x$ in $I$, where $x\neq c$,  
if  $|x - c| < \delta$, then $|f(x) - L| < \epsilon$.
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/Gw1k8A0chjg")
```

Mathematicians often enjoy writing ideas without using any words. Here
is the wordless definition of the limit:

$\displaystyle \lim_{x\rightarrow c} f(x) = L \iff$
$\forall \, \epsilon > 0, \exists \, \delta > 0 \; s.t. \;
0<|x - c| < \delta \longrightarrow |f(x) - L| < \epsilon$.

Note the order in which $\epsilon$ and $\delta$ are given. In the
definition, the $y$-tolerance $\epsilon$ is given *first* and then the
limit will exist **if**  we can find an $x$-tolerance $\delta$ that
works.


An example will help us understand this definition. Note that the
explanation is long, but it will take one through all steps necessary to
understand the ideas.\

### Example: Square root function

Show that $\displaystyle \lim_{x\rightarrow 4} \sqrt{x} = 2$.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/pdu3nRE8x1I")
```



Before we use the formal definition, let's try some numerical tolerances.  What if the $y$ tolerance is 0.5, or $\epsilon =0.5$?  How close to 4 does $x$ have to be so that $y$ is within 0.5 units of 2, i.e., $1.5 < y < 2.5$?  In this case, we can proceed as follows:
\begin{alignedat}{3}
1.5 &< y <& 2.5 \\
1.5 &< \sqrt{x} <& 2.5\\
1.5^2 &< x <& 2.5^2\\
2.25 &< x <& 6.25.
\end{alignedat}

So, what is the desired $x$ tolerance?  Remember, we want to find a symmetric interval of $x$ values, namely
$4 - \delta < x < 4 + \delta$.  The lower bound of $2.25$ is $1.75$ units from 4; the upper bound of 6.25 is 2.25 units from 4. We need the smaller of these two distances; we must have $\delta < 1.75$. See Figure \@ref(fig:chooseed).


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/aJ33_ItUoLk")
```

```{r chooseed, fig.cap="Epsilon delta visualization in Desmos. First set epsilon to determine the vertical tolerance, then find a corresponding delta for the horizontal tolerance which keeps the function within the horizontal band.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.desmos.com/calculator/nmihcrca3u")
```


Given the $y$ tolerance $\epsilon =0.5$, we have found an $x$ tolerance, $\delta < 1.75$, such that whenever $x$ is within $\delta$ units of 4, then $y$ is within $\epsilon$ units of 2.  That's what we were trying to find.
  
Let's try another value of $\epsilon$.

What if the $y$ tolerance is 0.01, i.e.,  $\epsilon =0.01$?  How close to 4 does $x$ have to be in order for $y$ to be within 0.01 units of 2 (or $1.99 < y < 2.01$)?  Again, we just square these values to get
$1.99^2 < x < 2.01^2$, or 
$$3.9601 < x < 4.0401.$$  
What is the desired $x$ tolerance?  In this case we must have $\delta < 0.0399$, which is the minimum distance from 4 of the two bounds given above.  

Note that in some sense, it looks like there are two tolerances (below 4 of 0.0399 units and above 4 of 0.0401 units).  However, we couldn't use the larger value of $0.0401$ for $\delta$ since then the interval for $x$ would be  $3.9599 < x < 4.0401$ resulting in $y$ values of $1.98995 < y < 2.01$ (which contains values NOT within 0.01 units of 2).

What we have so far: if $\epsilon =0.5$, then $\delta < 1.75$ and if $\epsilon = 0.01$, then $\delta < 0.0399$. A pattern is not easy to see, so we switch to general $\epsilon$ try to determine $\delta$ symbolically.  We start by assuming $y=\sqrt{x}$ is within $\epsilon$ units of 2:



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/zA_u1Hsd0Jg")
```



\begin{eqnarray*}
|y - 2| < \epsilon &\\
-\epsilon < y - 2 < \epsilon& \qquad \textrm{(Definition of absolute value)}\\
-\epsilon < \sqrt{x} - 2 < \epsilon  &\qquad (y=\sqrt{x})\\
2 - \epsilon < \sqrt{x} < 2+ \epsilon &\qquad \textrm{ (Add 2)}\\
(2 - \epsilon)^2 < x < (2+ \epsilon) ^2 &\qquad \textrm{ (Square all)}\\
4 - 4\epsilon + \epsilon^2 < x < 4 + 4\epsilon + \epsilon^2 &\qquad \textrm{ (Expand)}\\
4 - (4\epsilon - \epsilon^2) < x < 4 + (4\epsilon + \epsilon^2). &\qquad \textrm{ (Rewrite in the desired form)}
\end{eqnarray*}

The _"desired form"_ in the last step is ``$4-\textit{something} < x < 4 +\textit{something}$.''
Since we want this last interval to describe an $x$ tolerance around 4, we have that either $\delta < 4\epsilon - \epsilon^2$ or $\delta < 4\epsilon + \epsilon^2$, whichever is smaller: $$\delta < \min\{4\epsilon - \epsilon^2, 4\epsilon + \epsilon^2\}.$$  Since $\epsilon > 0$, the minimum is $\delta < 4\epsilon - \epsilon^2$.  That's the formula: given an $\epsilon$, set $\delta < 4\epsilon-\epsilon^2$. 

We can check this for our previous values.  If $\epsilon=0.5$, the formula gives
$\delta < 4(0.5) - (0.5)^2 = 1.75$ and when $\epsilon=0.01$, the formula gives $\delta < 4(0.01) - (0.01)^2 = 0.399$.


		
```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/Yt9l19jrnBU")
```


So given any $\epsilon >0$, set $\delta < 4\epsilon - \epsilon^2$. Then if $|x-4|<\delta$ (and $x\neq 4$), then $|f(x) - 2| < \epsilon$,  satisfying the definition of the limit.  We have shown formally (and finally!) that $\displaystyle \lim_{x\rightarrow 4} \sqrt{x} = 2 $.


The previous example was a little long in that we sampled a few specific
cases of $\epsilon$ before handling the general case. Normally this is
not done. The previous example is also a bit unsatisfying in that
$\sqrt{4}=2$; why work so hard to prove something so obvious? Many
$\epsilon$-$\delta$ proofs are long and difficult to do. In this
section, we will focus on examples where the answer is, frankly,
obvious, because the non–obvious examples are even harder. In the next
section we will learn some theorems that allow us to evaluate limits
*analytically*, that is, without using the $\epsilon$-$\delta$
definition.\

### Example: Square function

Show that $\displaystyle \lim_{x\rightarrow 2} x^2 = 4$.



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/QGqoq-xEXyk")
```


Let's do this example symbolically from the start.  Let $\epsilon > 0$ be given; we want $|y-4| < \epsilon$, i.e.,  $|x^2-4| < \epsilon$.  How do we find $\delta$ such that when $|x-2| < \delta$, we are guaranteed that $|x^2-4|<\epsilon$?


This is a bit trickier than the previous example, but let's start by noticing that 
$|x^2-4| = |x-2|\cdot|x+2|$.  Consider:
\begin{equation} |x^2-4| < \epsilon \longrightarrow |x-2|\cdot|x+2| < \epsilon \longrightarrow |x-2| < \frac{\epsilon}{|x+2|}.\label{eq:limit1}\end{equation} 
Could we not set $\displaystyle \delta = \frac{\epsilon}{|x+2|}$?  

We are close to an answer, but the catch is that $\delta$ must be a \textit{constant} value (so it can't contain $x$).  There is a way to work around this, but we do have to make an assumption.  Remember that $\epsilon$ is supposed to be a small number, which implies that $\delta$ will also be a small value.  In particular, we can (probably) assume that $\delta < 1$.  If this is true, then $|x-2| < \delta$ would imply that $|x-2| < 1$, giving $1 < x < 3$.  

Now, back to the fraction $\displaystyle \frac{\epsilon}{|x+2|}$.  If $1<x<3$, then $3<x+2<5$ (add 2 to all terms in the inequality).  Taking reciprocals, we have 
\begin{align}
\frac15 <& \frac1{|x+2|} < \frac 13 & \text{which implies}\notag\\
\frac15 <& \frac1{|x+2|} & \text{which implies}\notag\\
\frac\epsilon5<&\frac{\epsilon}{|x+2|}.\label{eq:limit2}
\end{align}

This suggests that we set 
$\displaystyle \delta < \frac{\epsilon}{5}$. To see why, let consider what follows when we assume $|x-2|<\delta$:

\small
\begin{align*}
|x - 2| &< \delta &\\
|x - 2| &< \frac{\epsilon}{5}&  \text{(Our choice of $\delta$)}\\
|x - 2|\cdot|x + 2| &< |x + 2|\cdot\frac{\epsilon}{5}&  \text{(Multiply by $|x+2|$)}\\
|x^2 - 4|&< |x + 2|\cdot\frac{\epsilon}{5}&  \text{(Combine left side)}\\
|x^2 - 4|&< |x + 2|\cdot\frac{\epsilon}{5}< |x + 2|\cdot\frac{\epsilon}{|x+2|}=\epsilon &  
\text{(As long as $\delta <1$)}
\end{align*}
\normalsize

We have arrived at $|x^2 - 4|<\epsilon$ as desired.  Note again, in order to make this happen we needed $\delta$ to first be less than 1.  That is a safe assumption; we want $\epsilon$ to be arbitrarily small, forcing $\delta$ to also be small. 

We have also picked $\delta$ to be smaller than ``necessary.'' We could get by with a slightly larger $\delta$, as shown in Figure \@ref(fig:limit_eover5). The dashed outer lines show the boundaries defined by our choice of $\epsilon$. The dotted inner lines show the boundaries defined by setting $\delta = \epsilon/5$. Note how these dotted lines are within the dashed lines. That is perfectly fine; by choosing $x$ within the dotted lines we are guaranteed that $f(x)$ will be within $\epsilon$ of 4.


		
```{r squareed, fig.cap="Epsilon delta visualization in Desmos. First set epsilon to determine the vertical tolerance, then find a corresponding delta for the horizontal tolerance which keeps the function within the horizontal band.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.desmos.com/calculator/w7o708dap6")
```


In summary, given $\epsilon > 0$, set $\delta=\epsilon/5$.  Then $|x - 2| < \delta$ implies 
$|x^2 - 4|< \epsilon$ (i.e. $|y - 4|< \epsilon$) as desired.  This shows that $\displaystyle \lim_{x\rightarrow 2} x^2 = 4 $. Figure \@ref(fig:limit_eover5) gives a visualization of this; by restricting $x$ to values within $\delta = \epsilon/5$ of 2, we see that $f(x)$ is within $\epsilon$ of $4$.



Make note of the general pattern exhibited in these last two examples.
In some sense, each starts out "backwards." That is, while we want to

1.  start with $|x-c|<\delta$ and conclude that

2.  $|f(x)-L|<\epsilon$,

we actually start by assuming

1.  $|f(x)-L|<\epsilon$, then perform some algebraic manipulations to
    give an inequality of the form

2.  $|x-c|<$ *something*.

When we have properly done this, the *something* on the "greater than"
side of the inequality becomes our $\delta$. We can refer to this as the
"scratch–work" phase of our proof. Once we have $\delta$, we can
formally start with $|x-c|<\delta$ and use algebraic manipulations to
conclude that $|f(x)-L|<\epsilon$, usually by using the same steps of
our "scratch–work" in reverse order.

We highlight this process in the following example.

### Example: Polynomial

We illustrate evaluating limits once more.



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/--_Rq2GX9IY")
```

Prove that $\displaystyle \lim_{x\rightarrow 1}(x^3-2x) = -1$.

We start our scratch--work by considering $|f(x) - (-1)| < \epsilon$:
\begin{align}
|f(x)-(-1)| &< \epsilon \notag\\
|x^3-2x + 1|&< \epsilon & \text{(Now factor)}\notag\\
|(x-1)(x^2+x-1)|&< \epsilon \notag\\
|x-1| &<\frac{\epsilon}{|x^2+x-1|}.\label{eq:lim4}
\end{align}
We are at the phase of saying that $|x-1|<$ \textit{something}, where \textit{something}$=\epsilon/|x^2+x-1|$. We want to turn that \textit{something} into $\delta$.

Since $x$ is approaching 1, we are safe to assume that $x$ is between 0 and 2. So
\begin{align*}
0&< x<2  & \\
0&< x^2<4.&\text{(squared each term)}
\end{align*}
Since $0<x<2$, we can add $0$, $x$ and $2$, respectively, to each part of the inequality and maintain the inequality.
\begin{align*}
0&< x^2+x<6 &\\
-1&< x^2+x-1<5.&\text{(subtracted 1 from each part)}
\end{align*}

In Equation \eqref{eq:lim4}, we wanted $|x-1|<\epsilon/|x^2+x-1|$. The above shows that given any $x$ in $[0,2]$, we know that 
\begin{align}
x^2+x-1 &< 5 &\text{which implies that}\notag\\
\frac15 &< \frac{1}{x^2+x-1} &\text{which implies that}\notag\\
\frac{\epsilon}5 &< \frac{\epsilon}{x^2+x-1}.\label{eq:lim4b}
\end{align}
 So we set $\delta < \epsilon/5$. This ends our scratch--work, and we begin the formal proof (which also helps us understand why this was a good choice of $\delta$).

Given $\epsilon$, let $\delta < \epsilon/5$. We want to show that when $|x-1|<\delta$, then $|(x^3-2x)-(-1)|<\epsilon$. We start with $|x-1|<\delta$:
\begin{align*}
|x-1| &< \delta \\
|x-1| &< \frac{\epsilon}5\\
|x-1| &< \frac\epsilon5 < \frac{\epsilon}{|x^2+x-1|} & \text{(for $x$ near 1, from Equation \eqref{eq:lim4b})}\\
|x-1|\cdot |x^2+x-1| &< \epsilon\\
|x^3-2x+1| &< \epsilon\\
|(x^3-2x)-(-1)| &<\epsilon,
\end{align*}
which is what we wanted to show. Thus $\displaystyle \lim_{x\to 1}(x^3-2x) = -1$.


### Example: Exponential

Prove that $\displaystyle \lim_{x\rightarrow 0} e^x = 1$.

Symbolically, we want to take the equation $|e^x - 1| < \epsilon$ and unravel it to the form $|x-0| < \delta$.  Here is our scratch--work:
\begin{eqnarray*}
|e^x - 1| < \epsilon&\\
-\epsilon < e^x - 1 < \epsilon& \qquad \textrm{(Definition of absolute value)}\\
1-\epsilon < e^x < 1+\epsilon & \qquad \textrm{(Add 1)}\\
\ln(1-\epsilon) < x < \ln(1+\epsilon) & \qquad \textrm{(Take natural logs)}\\
\end{eqnarray*}
Making the safe assumption that $\epsilon<1$ ensures the last inequality is valid (i.e., so that $\ln (1-\epsilon)$ is defined). We can then set $\delta$ to be the minimum of $|\ln(1-\epsilon)|$ and $\ln(1+\epsilon)$; i.e., 
$$\delta = \min\{|\ln(1-\epsilon)|, \ln(1+\epsilon)\} = \ln(1+\epsilon).$$  


Now, we work through the actual the proof:

\begin{align*}
|x - 0|&<\delta\\
-\delta &< x < \delta &  \textrm{(Definition of absolute value)}\\
-\ln(1+\epsilon) &< x < \ln(1+\epsilon). &\\  
\ln(1-\epsilon) &< x < \ln(1+\epsilon). & \text{(since $\ln(1-\epsilon) < -\ln(1+\epsilon)$)}\\ 
\end{align*}
The above line is true by our choice of $\delta$ and by the fact that since $|\ln(1-\epsilon)|>\ln(1+\epsilon)$ and $\ln(1-\epsilon)<0$, we know $\ln(1-\epsilon) < -\ln(1+\epsilon )$.
\begin{align*}
1-\epsilon &< e^x < 1+\epsilon &  \textrm{(Exponentiate)}\\
-\epsilon &< e^x - 1 < \epsilon &  \textrm{(Subtract 1)}\\
\end{align*}

In summary, given $\epsilon > 0$, let $\delta = \ln(1+\epsilon)$. Then $|x - 0| < \delta$ implies $|e^x - 1|< \epsilon$ as desired.  We have shown that $\displaystyle \lim_{x\rightarrow 0} e^x = 1$.


We note that we could actually show that
$\lim_{x\rightarrow c} e^x = e^c$ for any constant $c$. We do this by
factoring out $e^c$ from both sides, leaving us to show
$\lim_{x\rightarrow c} e^{x-c} = 1$ instead. By using the substitution
$u=x-c$, this reduces to showing $\lim_{u\rightarrow 0} e^u = 1$ which
we just did in the last example. As an added benefit, this shows that in
fact the function $f(x)=e^x$ is *continuous* at all values of $x$, an
important concept we will define in Section \@ref(sec:continuity).


### Conclusion {-}


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/BapEm7_GQnM")
```



This formal definition of the limit is not an easy concept grasp. Our
examples are actually "easy" examples, using "simple" functions like
polynomials, square–roots and exponentials. It is very difficult to
prove, using the techniques given above, that
$\displaystyle \lim_{x\to 0}(\sin x)/x = 1$, as we approximated in the previous
section.

There is hope. The next section shows how one can evaluate complicated
limits using certain basic limits as building blocks. While limits are
an incredibly important part of calculus (and hence much of higher
mathematics), rarely are limits evaluated using the definition. Rather,
the techniques of the following section are employed.

## Finding Limits Analytically {#sec:limit_analytically}


In Section \@ref(sec:limitintro) we explored the concept of the limit
without a strict definition, meaning we could only make approximations.
In the previous section we gave the definition of the limit and
demonstrated how to use it to verify our approximations were correct.
Thus far, our method of finding a limit is 1) make a really good
approximation either graphically or numerically, and 2) verify our
approximation is correct using a $\epsilon$-$\delta$ proof.

Recognizing that $\epsilon$-$\delta$ proofs are cumbersome, this section
gives a series of theorems which allow us to find limits much more
quickly and intuitively.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/idVyYI9XH-A")
```


Suppose that $\lim_{x\to 2} f(x)=2$ and $\lim_{x\to 2} g(x) = 3$. What
is $\lim_{x\to 2}(f(x)+g(x))$? Intuition tells us that the limit should
be 5, as we expect limits to behave in a nice way. The following theorem
states that already established limits do behave nicely.

>
```{theorem, label="limitalgebra", name="Basic Limit Properties", echo=TRUE}
Let $b$, $c$, $L$ and $K$ be real numbers, let $n$ be a positive integer, and let $f$ and $g$ be functions defined on an open interval $I$ containing $c$ with the following limits: \index{limit!properties}
$$\lim_{x\to c}f(x) = L \text{\ and\ } \lim_{x\to c} g(x) = K.$$
The following limits hold.
>
1. Constants: $\displaystyle \lim_{x\to c} b = b$
1. Identity:						$\displaystyle \lim_{x\to c} x = c$
1. Sums/Differences: $\displaystyle \lim_{x\to c}(f(x)\pm g(x)) = L\pm K$
1. Scalar Multiples:	$\displaystyle \lim_{x\to c} b\cdot f(x) = bL$
1. Products:	$\displaystyle \lim_{x\to c} f(x)\cdot g(x) = LK$
1. Quotients: $\displaystyle \lim_{x\to c} f(x)/g(x) = L/K$, ($K\neq 0)$
1. Powers: 	$\displaystyle \lim_{x\to c} f(x)^n = L^n$
1. Roots:		$\displaystyle \lim_{x\to c} \sqrt[n]{f(x)} = \sqrt[n]{L}$ (If $n$ is even then require $f(x)\geq 0$ on $I$.)
1. Compositions: Adjust our previously given limit situation to: $$\lim_{x\to c}f(x) = L,\ \lim_{x\to L} g(x) = K \text{ and } g(L)=K .$$ Then $\displaystyle \lim_{x\to c}g(f(x)) = K$.

```


We make a note about Property \#8: when $n$ is even, $L$ must be greater
than 0. If $n$ is odd, then the statement is true for all $L$.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/5P2Zac-_Xhg")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/lNbLZ2BR1gQ")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/2BNMJMxxAsM")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/l0XDW-reRlI")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/06IBnq9hXY8")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/9wKX2JgVHXA")
```



### Example: Using basic limit properties.

We apply the theorem to an example.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/8x42kGfu9ts")
```


Let $$\lim_{x\to 2} f(x)=2,\quad\lim_{x\to 2} g(x) = 3\quad \text{ and }\quad p(x) = 3x^2-5x+7.$$ Find the following limits:


1.		$\displaystyle \lim_{x\to 2} \big(f(x) + g(x)\big)$
1.		$\displaystyle \lim_{x\to 2} \big(5f(x) + g(x)^2\big)$
1.		$\displaystyle \lim_{x\to 2} p(x)$

Solutions:

1.		Using the Sum/Difference rule, we know that $\displaystyle \lim_{x\to 2} \big(f(x) + g(x)\big) = 2+3 =5$.
1.		Using the Scalar Multiple and Sum/Difference rules, we find that $\displaystyle \lim_{x\to 2} \big(5f(x) + g(x)^2\big) = 5\cdot 2 + 3^2 = 19.$
1.		Here we combine the Power, Scalar Multiple, Sum/Difference and Constant Rules. We show quite a few steps, but in general these can be omitted:
				\begin{align*}
				\lim_{x\to 2} p(x) &= \lim_{x\to 2} (3x^2-5x+7) \\
				&= \lim_{x\to 2} 3x^2-\lim_{x\to 2} 5x+\lim_{x\to 2}7 \\
				 &= 3\cdot 2^2 - 5\cdot 2+7 \\
				 &= 9
				\end{align*}

### Limits of polynomial and rational functions {-}

Part 3 of the previous example demonstrates how the limit of a quadratic
polynomial can be determined using the properties of Theorem
\@ref(thm:limitalgebra). Not only that, recognize that
$$\lim_{x\to 2} p(x) = 9 = p(2)$$ i.e., the limit at 2 was found just
by plugging 2 into the function. This holds true for all polynomials,
and also for rational functions (which are quotients of polynomials), as
stated in the following theorem.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/NGoUHZESPso")
```


>
```{theorem, label="polyrat", title="Limits of Polynomial and Rational Functions", echo=TRUE}
Let $p(x)$ and $q(x)$ be polynomials and $c$ a real number. Then:
>
1.	$\displaystyle \lim_{x\to c} p(x) = p(c)$
1. $\displaystyle \lim_{x\to c} \frac{p(x)}{q(x)} = \frac{p(c)}{q(c)}$, where $q(c) \neq 0$.

```

### Example: Limit of a rational function {#ex:limrat}

Using Theorem \@ref(thm:polyrat), find $$\lim_{x\to -1} \frac{3x^2-5x+1}{x^4-x^2+3}.$$

Using Theorem \@ref(thm:polyrat), we can quickly state that 
	\begin{align*} 
	\lim_{x\to -1}\frac{3x^2-5x+1}{x^4-x^2+3} &= \frac{3(-1)^2-5(-1)+1}{(-1)^4-(-1)^2+3} \\
												&= \frac{9}{3} =3.
	\end{align*}


------

It was likely frustrating in Section \@ref(sec:limitdef) to do a lot of
work to prove that $$\lim_{x\to 2} x^2 = 4$$ as it seemed fairly
obvious. The previous theorems state that many functions behave in such
an "obvious" fashion, as demonstrated by the rational function in
Example \@ref(ex:limrat).

### Example: Evaluating limits using algebra {#ex:limitonept}
Evaluate the following limit: $$\lim_{x\to 1}\frac{x^2-1}{x-1}.$$

We begin by attempting to apply Theorem \@ref(thm:polyrat) and substituting 1 for $x$ in the quotient. This gives:
		$$\lim_{x\to 1}\frac{x^2-1}{x-1} = \frac{1^2-1}{1-1} \to \frac{0}{0},$$ an indeterminate form. We cannot apply the theorem.


```{r limitxplus1, fig.cap="The function for Example \\@ref(ex:limitonept)", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/bbheqmd3")
```

		
\mfigure{.6}{Graphing $f$ in Example \@ref(ex_limit_onept) to understand a limit.}{fig:limitxplus1}{figures/fig_LimitXplus1}
		
		By graphing the function, as in Figure \@ref(fig:limitxplus1), we see that the function seems to be linear, implying that the limit should be easy to evaluate. Recognize that the numerator of our quotient can be factored:
		$$\frac{x^2-1}{x-1} = \frac{(x-1)(x+1)}{x-1}.$$
		The function is not defined when $x=1$, but for all other $x$, $$\frac{x^2-1}{x-1} = \frac{(x-1)(x+1)}{x-1} = x+1.$$
		Clearly $\displaystyle \lim_{x\to 1}x+1 = 2$. Recall that when considering limits, we are not concerned with the value of the function at 1, only the value the function approaches as $x$ approaches 1. Since $(x^2-1)/(x-1)$ and $x+1$ are the same at all points except $x=1$, they both approach the same value as $x$ approaches 1. Therefore we can conclude that $$\lim_{x\to 1}\frac{x^2-1}{x-1}=2.$$



### Functions equal everywhere except one point {-}

The key to the above example is that the functions $y=(x^2-1)/(x-1)$ and
$y=x+1$ are identical except at $x=1$. Since limits describe a value the
function is approaching, not the value the function actually attains,
the limits of the two functions are always equal.

>
```{theorem, label="limitallbut1", name="Limits of Functions Equal At All But One Point", echo=TRUE}
Let $g(x) = f(x)$ for all $x$ in an open interval, except possibly at $c$, and let $\displaystyle \lim_{x\to c} g(x) = L$ for some real number $L$. Then $$\lim_{x\to c}f(x) = L.$$
```

The Fundamental Theorem of Algebra tells us that when dealing with a
rational function of the form $g(x)/f(x)$ and directly evaluating the
limit $\displaystyle \lim_{x\to c} \frac{g(x)}{f(x)}$ returns "0/0", then $(x-c)$
is a factor of both $g(x)$ and $f(x)$. One can then use algebra to
factor this term out, cancel, then apply Theorem \@ref(thm:limitallbut1).
We demonstrate this once more.

### Example: Evaluating a limit using Theorem \@ref(thm:limitallbut1) {#ex:limitallbut1}

Evaluate $\displaystyle \lim_{x\to 3} \frac{x^3-2 x^2-5 x+6}{2 x^3+3 x^2-32 x+15}$.



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/MSckoYqdKH4")
```


We attempt to apply Theorem \@ref(thm:polyrat) by substituting 3 for $x$. This returns the familiar indeterminate form of "0/0".

Since the numerator and denominator are each polynomials, we know that $(x-3)$ is factor of each. Using whatever method is most comfortable to you, factor out $(x-3)$ from each (using polynomial division, synthetic division, a computer algebra system, etc.). We find that $$\frac{x^3-2 x^2-5 x+6}{2 x^3+3 x^2-32 x+15} = \frac{(x-3)(x^2+x-2)}{(x-3)(2 x^2+9 x-5)}.$$ We can cancel the $(x-3)$ terms as long as $x\neq 3$. Using Theorem \@ref(thm:limitallbut1) we conclude:
		\begin{align*}
		\lim_{x\to 3} \frac{x^3-2 x^2-5 x+6}{2 x^3+3 x^2-32 x+15} &= \lim_{x\to 3}\frac{(x-3)(x^2+x-2)}{(x-3)(2 x^2+9 x-5)} \\
																															&=	\lim_{x\to 3} \frac{(x^2+x-2)}{(2 x^2+9 x-5)}\\
																															&= \frac{10}{40} = \frac14.
		\end{align*}

----


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/vOW92eipOu4")
```

### Other limits by substitution {-}

Polynomial and rational functions are not the only functions to behave
in such a predictable way. The following theorem gives a list of
functions whose behavior is particularly "nice" in terms of limits. In
the next section, we will give a formal name to these functions that
behave "nicely."


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/S0VmjueQn5k")
```


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/L2ohy96L-F4")
```


>
```{theorem, label="limcontinuous", name="Special Limits", echo=TRUE}
Let $c$ be a real number in the domain of the given function and let $n$ be a positive integer. The following limits hold: 
>
1.		$\displaystyle \lim_{x\to c} \sin x = \sin c$
1.		$\displaystyle \lim_{x\to c} \cos x = \cos c$
1.		$\displaystyle \lim_{x\to c} \tan x = \tan c$
1.		$\displaystyle \lim_{x\to c} \csc x = \csc c$
1.		$\displaystyle \lim_{x\to c} \sec x = \sec c$
1.		$\displaystyle \lim_{x\to c} \cot x = \cot c$
1.		$\displaystyle \lim_{x\to c} a^x = a^c$ ($a>0$)
1.		$\displaystyle \lim_{x\to c} \ln x = \ln c$
1.		$\displaystyle \lim_{x\to c} \sqrt[n]{x} = \sqrt[n]{c}$
>
```


### Example: Evaluating limits analytically {#ex:limit1}

Evaluate the following limits. 

1.		$\displaystyle \lim_{x\to \pi} \cos x$
1.		$\displaystyle \lim_{x\to 3} (\sec^2x - \tan^2 x)$
1.		$\displaystyle \lim_{x\to \pi/2} \cos x\sin x$
1.		$\displaystyle \lim_{x\to 1} e^{\ln x}$
1.		$\displaystyle \lim_{x\to 0} \frac{\sin x}{x}$

Solutions:

1.		This is a straightforward application of Theorem \@ref(thm:lim_continuous). $\displaystyle \lim_{x\to \pi} \cos x = \cos \pi = -1$.
1.		We can approach this in at least two ways. First, by directly applying Theorem \@ref(thm:lim_continuous), we have:
				$$\lim_{x\to 3} (\sec^2x - \tan^2 x) = \sec^23-\tan^23.$$ Using the Pythagorean Theorem, this last expression is 1; therefore $$\lim_{x\to 3} (\sec^2x - \tan^2 x) = 1.$$ We can also use the Pythagorean Theorem from the start. $$\lim_{x\to 3} (\sec^2x - \tan^2 x) = \lim_{x\to 3} 1 = 1,$$ using the Constant limit rule. Either way, we find the limit is 1.
1.		Applying the Product limit rule of Theorem \@ref(thm:limit_algebra) and Theorem \@ref(thm:lim_continuous) gives $$\displaystyle \lim_{x\to \pi/2} \cos x\sin x = \cos (\pi/2)\sin(\pi/2) = 0\cdot 1 = 0.$$
1.		Again, we can approach this in two ways. First, we can use the exponential/logarithmic identity that $e^{\ln x} = x$ and evaluate $\displaystyle \lim_{x\to 1} e^{\ln x} = \lim_{x\to 1} x = 1.$ 
We can also use the limit Composition Rule of Theorem \@ref(thm:limit_algebra). Using Theorem \@ref(thm:lim_continuous), we have $\displaystyle \lim_{x\to 1}\ln x = \ln 1 = 0$ and $\displaystyle\lim_{x\to 0} e^x= e^0=1$, satisfying the conditions of the Composition Rule. Applying this rule, $$\displaystyle \lim_{x\to 1} e^{\ln x} = \lim_{x\to 0} e^x = e^0 = 1.$$ Both approaches are valid, giving the same result.

1.		We encountered this limit in Section \@ref(sec:limit_intro). Applying our theorems, we attempt to find the limit as $$\lim_{x\to 0}\frac{\sin x}{x}\rightarrow \frac{\sin 0}{0} \rightarrow \frac{0}{0}.$$ This, of course, violates a condition of Theorem \@ref(thm:limitalgebra), as the limit of the denominator is not allowed to be 0. Therefore, we are still unable to evaluate this limit with tools we currently have at hand.

### Squeeze Theorem {-}

The section could have been titled "Using Known Limits to Find Unknown
Limits." By knowing certain limits of functions, we can find limits
involving sums, products, powers, etc., of these functions. We further
the development of such comparative tools with the Squeeze Theorem, a
clever and intuitive way to find the value of some limits.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/8Tv-GRQdAVA")
```


Before stating this theorem formally, suppose we have functions $f$, $g$
and $h$ where $g$ always takes on values between $f$ and $h$; that is,
for all $x$ in an interval, $$f(x) \leq g(x) \leq h(x).$$ If $f$ and $h$
have the same limit at $c$, and $g$ is always "squeezed" between them,
then $g$ must have the same limit as well. That is what the Squeeze
Theorem states.

>
```{theorem, name="Squeeze Theorem", label="sqz", echo=TRUE}
Let $f$, $g$ and $h$ be functions on an open interval $I$ containing $c$ such that for all $x$ in $I$, $f(x)\leq g(x) \leq h(x).$ 
>
If $$\lim_{x\to c} f(x) = L = \lim_{x\to c} h(x),$$ then $$\lim_{x\to c} g(x) = L.$$
>
```


It can take some work to figure out appropriate functions by which to
"squeeze" a given function. However, that is generally the only place
where work is necessary; the theorem makes the "evaluating the limit
part" very simple.

We use the Squeeze Theorem in the following example to finally prove
that $\displaystyle \lim_{x\to 0} \frac{\sin x}{x} = 1$.\


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/pgjv3ojtXh4")
```

		
```{r sqzsinxoverx, fig.cap="Squeeze Theorem applied to $\\sin(x)/x$.", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.desmos.com/calculator/291p7twtnd")
```


Two notes about the previous example are worth mentioning. First, one
might be discouraged by this application, thinking "I would *never* have
come up with that on my own. This is too hard!" Don’t be discouraged;
within this text we will guide you in your use of the Squeeze Theorem.
As one gains mathematical maturity, clever proofs like this are easier
and easier to create.

Second, this limit tells us more than just that as $x$ approaches 0,
$\sin(x)/x$ approaches 1. Both $x$ and $\sin x$ are approaching 0, but
the *ratio* of $x$ and $\sin x$ approaches 1, meaning that they are
approaching 0 in essentially the same way. Another way of viewing this
is: for small $x$, the functions $y=x$ and $y=\sin x$ are essentially
indistinguishable.\



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/Wd464IIls5Y")
```


We include this special limit, along with three others, in the following
theorem.

>
```{theorem, label="speciallimits", name="Special Limits", echo=TRUE}
These special limits are useful in finding limits of difference quotients.
>
1. $\displaystyle \lim_{x\to 0} \frac{\sin x}{x} = 1$
1. $\displaystyle \lim_{x\to 0} \frac{\cos x-1}{x} = 0$
1. $\displaystyle \lim_{x\to 0} (1+x)^\frac1x = e$
1. $\displaystyle \lim_{x\to 0} \frac{e^x-1}{x} = 1$
>
```

A short word on how to interpret the latter three limits. We know that
as $x$ goes to 0, $\cos x$ goes to 1. So, in the second limit, both the
numerator and denominator are approaching 0. However, since the limit is
0, we can interpret this as saying that "$\cos x$ is approaching 1
faster than $x$ is approaching 0."

In the third limit, inside the parentheses we have an expression that is
approaching 1 (though never equaling 1), and we know that 1 raised to
any power is still 1. At the same time, the power is growing toward
infinity. What happens to a number near 1 raised to a very large power?
In this particular case, the result approaches Euler’s number, $e$,
approximately $2.718.$

In the fourth limit, we see that as $x\to 0$, $e^x$ approaches 1 "just
as fast" as $x\to 0$, resulting in a limit of 1.\





We end this section by revisiting a limit first seen in Section
\@ref(sec:limitintro), a limit of a difference quotient. Let
$f(x) = -1.5x^2+11.5x$; we approximated the limit
$\displaystyle \lim_{h\to 0}\frac{f(1+h)-f(1)}{h}\approx 8.5.$ We formally
evaluate this limit in the following example.\

### Example: Evaluating the limit of a difference quotient {#ex:limitdiffquot}

Let $f(x) = -1.5x^2+11.5x$; find $\displaystyle \lim_{h\to 0}\frac{f(1+h)-f(1)}{h}.$

Since $f$ is a polynomial, our first attempt should be to employ Theorem \@ref(thm:polyrat) and substitute 0 for $h$. However, we see that this gives us "$0/0$."
 Knowing that we have a rational function hints that some algebra will help. Consider the following steps:
		\begin{align*}
		\lim_{h\to 0}\frac{f(1+h)-f(1)}{h} 	&= 	\lim_{h\to 0}\frac{-1.5(1+h)^2 + 11.5(1+h) - \left(-1.5(1)^2+11.5(1)\right)}{h} \\
																				&=	\lim_{h\to 0}\frac{-1.5(1+2h+h^2) + 11.5+11.5h - 10}{h}\\
																				&=	\lim_{h\to 0}\frac{-1.5h^2 +8.5h}{h}\\
																				&= 	\lim_{h\to 0}\frac{h(-1.5h+8.5)}h\\
																				&=	\lim_{h\to 0}(-1.5h+8.5)  \\
																				&= 	8.5 
		\end{align*}																		
This matches our previous approximation.


### Conclusion {-}

This section contains several valuable tools for evaluating limits. One
of the main results of this section is Theorem \@ref(thm:limcontinuous);
it states that many functions that we use regularly behave in a very
nice, predictable way. In Section \@ref(sec:continuity) we give a name to
this nice behavior; we label such functions as *continuous.* Defining
that term will require us to look again at what a limit is and what
causes limits to not exist.



## One Sided Limits {#sec:limit_continuity}

We introduced the concept of a limit gently, approximating their values
graphically and numerically. Next came the rigorous definition of the
limit, along with an admittedly tedious method for evaluating them. The
previous section gave us tools (which we call theorems) that allow us to
compute limits with greater ease. Chief among the results were the facts
that polynomials and rational, trigonometric, exponential and
logarithmic functions (and their sums, products, etc.) all behave
"nicely." In this section we rigorously define what we mean by "nicely."

In Section \@ref(sec:limitintro) we saw three ways in which limits of
functions failed to exist:

1.  The function approached different values from the left and right,

2.  The function grows without bound, and

3.  The function oscillates.

In this section we explore in depth the concepts behind \#1 by
introducing the *one-sided limit*. We begin with formal definitions that
are very similar to the definition of the limit given in Section
\@ref(sec:limitdef), but the notation is slightly different and
"$x\neq c$" is replaced with either "$x<c$" or "$x>c$."


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/VU8lUocFAfE")
```

>
```{definition, label="onesidedlimit", name="One Sided Limits: Left- and Right-Hand Limits", echo=TRUE}

>
Left-Hand Limit
>
: Let $f$ be a function defined on $(a,c)$ for some $a<c$ and let $L$ be a real number. The _limit of $f(x)$, as $x$ approaches $c$ from the left, is $L$_, or, the left-hand limit of $f$ at $c$ is $L$, denoted by  
$$\displaystyle \lim_{x\rightarrow c^-} f(x) = L,$$
means  given any $\epsilon > 0$, there exists $\delta > 0$ such that for all $a<x<c$,  
if  $|x - c| < \delta$, then $|f(x) - L| < \epsilon$.
>
Right-Hand Limit
>
: Let $f$ be a function defined on $(c,b)$ for some $b>c$ and let $L$ be a real number. 
>
The limit of $f(x)$, as $x$ approaches $c$ from the right, is $L$_, or, the right-hand limit of $f$ at $c$ is $L$, denoted by $$\displaystyle \lim_{x\rightarrow c^+} f(x) = L,$$ means  given any $\epsilon > 0$, there exists $\delta > 0$ such that for all $c<x<b$,  
if  $|x - c| < \delta$, then $|f(x) - L| < \epsilon$.
>
```


Practically speaking, when evaluating a left-hand limit, we consider
only values of $x$ "to the left of $c$," i.e., where $x<c$. The
admittedly imperfect notation $x\to c^-$ is used to imply that we look
at values of $x$ to the left of $c$. The notation has nothing to do with
positive or negative values of either $x$ or $c$. A similar statement
holds for evaluating right-hand limits; there we consider only values of
$x$ to the right of $c$, i.e., $x>c$. We can use the theorems from
previous sections to help us evaluate these limits; we just restrict our
view to one side of $c$.


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/zQ5n78F49Bo")
```


We practice evaluating left- and right-hand limits through a series of
examples.

### Example: Evaluating one sided limits {#ex:onesidea}

Let $\displaystyle f(x) = \begin{cases} x & 0\leq x\leq 1 \\ 3-x & 1<x<2\end{cases},$ as shown in Figure \@ref(fig:onesided1). Find each of the following: 


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/NdBPwaP4Xkk")
```

		
```{r onesided1, fig.cap="The function for Example \\@ref(ex:onesidea)", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/t9dpjyng")
```


1.	$\displaystyle \lim_{x\to 1^-} f(x)$
1.	$\displaystyle \lim_{x\to 1^+} f(x)$
1.	$\displaystyle \lim_{x\to 1} f(x)$
1.	$\displaystyle f(1)$
1.	$\displaystyle \lim_{x\to 0^+} f(x)$
1.	$f(0)$
1.	$\displaystyle \lim_{x\to 2^-} f(x)$
1.	$f(2)$

For these problems, the visual aid of the graph is likely more effective in evaluating the limits than using $f$ itself. Therefore we will refer often to the graph.

1.	As $x$ goes to 1 \textit{from the left}, we see that $f(x)$ is approaching the value of 1. Therefore $\displaystyle \lim_{x\to 1^-} f(x) =1.$
1.	As $x$ goes to 1 \textit{from the right}, we see that $f(x)$ is approaching the value of 2. Recall that it does not matter that there is an ``open circle'' there; we are evaluating a limit, not the value of the function. Therefore $\displaystyle \lim_{x\to 1^+} f(x)=2$.
1.	\textit{The} limit of $f$ as $x$ approaches 1 does not exist, as discussed in the first section. The function does not approach one particular value, but two different values from the left and the right.
1.	Using the definition and by looking at the graph we see that $f(1) = 1$.
1.	As $x$ goes to 0 from the right, we see that $f(x)$ is also approaching 0. Therefore $\displaystyle \lim_{x\to 0^+} f(x)=0$. Note we cannot consider a left-hand limit at 0 as $f$ is not defined for values of $x<0$.
1.	Using the definition and the graph, $f(0) = 0$.
1.	As $x$ goes to 2 from the left, we see that $f(x)$ is approaching the value of 1. Therefore $\displaystyle \lim_{x\to 2^-} f(x)=1.$
1.	The graph and the definition of the function show that $f(2)$ is not defined.
			
----



Note how the left and right-hand limits were different at $x=1$. This,
of course, causes *the* limit to not exist. The following theorem states
what is fairly intuitive: *the* limit exists precisely when the left and
right-hand limits are equal.

>
```{theorem, label="leftrightlimits", name="Limits and One Sided Limits", echo=TRUE}
Let $f$ be a function defined on an open interval $I$ containing $c$. Then $$\lim_{x\to c}f(x) = L$$ if, and only if, $$\lim_{x\to c^-}f(x) = L \quad \text{ and } \quad \lim_{x\to c^+}f(x) = L.$$
```  


The phrase "if, and only if" means the two statements are *equivalent*:
they are either both true or both false. If the limit equals $L$, then
the left and right hand limits both equal $L$. If the limit is not equal
to $L$, then at least one of the left and right-hand limits is not equal
to $L$ (it may not even exist).

One thing to consider in Examples \@ref(ex:onesidea) – \@ref(ex:onesided) is
that the value of the function may/may not be equal to the value(s) of
its left/right-hand limits, even when these limits agree.


### Example: Evaluating limits of a piecewise--defined function {#ex:onesideb}
Let $f(x) = \begin{cases} 2-x & 0<x<1 \\ (x-2)^2 & 1<x<2 \end{cases}.$  Evaluate the following. 


1. $\displaystyle \lim_{x\to 1^-} f(x)$
1. $\displaystyle \lim_{x\to 1^+} f(x)$
1. $\displaystyle \lim_{x\to 1} f(x)$
1. $\displaystyle f(1)$
1. $\displaystyle \lim_{x\to 0^+} f(x)$
1. $f(0)$
1. $\displaystyle \lim_{x\to 2^-} f(x)$
1. $f(2)$



```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/vE_7FG2h_LU")
```


In this example, we evaluate each expression using just the definition of $f$, without using a graph as we did in the previous example.


1. As $x$ approaches 1 from the left, we consider a limit where all $x$-values are less than 1. This means we use the $2-x$ piece of the piecewise function $f$ as the domain for that piece is $(0,1)$. As the $x$-values near 1, $2-x$ approaches 1; that is, $f(x)$ approaches 1. Therefore $\displaystyle \lim_{x\to 1^-} f(x)=1.$
1. As $x$ approaches 1 from the right, we consider a limit where all $x$-values are greater than 1. This means we use the $(x-2)^2$ piece of $f$ as the domain for that piece is $(1,2)$. As the $x$-values near 1, $(x-2)^2$ approaches 1; that is, we see that again $f(x)$ approaches 1. Therefore $\displaystyle \lim_{x\to 1+} f(x)=1$.
1. \textit{The} limit of $f$ as $x$ approaches 1 exists and is 1, as $f$ approaches 1 from both the right and left. Therefore $\displaystyle \lim_{x\to 1} f(x)=1$.
1. Neither piece of $f$ is defined for the $x$-value of 1; in other words, 1 is not in the domain of $f$. Therefore $f(1)$ is not defined.  
1. As $x$ approaches  0 from the right, we consider a limit where all $x$-values are greater than 0. This means we use the $2-x$ piece of $f$. As the $x$-values near 0, $2-x$ approaches 2; that is, $f(x)$ approaches 2. So $\displaystyle \lim_{x\to 0^+} f(x)=2$.
1. $f(0)$  is not defined as $0$ is not in the domain of $f$.
1. As $x$ approaches  2 from the left, we consider a limit where all $x$-values are less than 2. This means we use the $(x-2)^2$ piece of $f$. As the $x$-values near 2, $(x-2)^2$ nears 0; that is, $f(x)$ approaches 0. So $\displaystyle \lim_{x\to 2^-} f(x)=0$.
1. $f(2)$  is not defined as 2 is not in the domain of $f$.

We can confirm our analytic result by consulting the graph of $f$ shown in Figure \ref{fig:onesidedb}.	Note the open circles on the graph at $x=0$, $1$ and $2$, where $f$ is not defined.	

		
```{r onesidedb, fig.cap="The function for Example \\@ref(ex:onesideb)", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/y3zaqrhv")
```



### Example: Evaluating limits of a piecewise--defined function {#ex:onesidec}

Let $f(x) = \begin{cases} (x-1)^2 & 0\leq x\leq 2, x\neq 1\\ 1 & x=1\end{cases},$ as shown in Figure \@ref(fig:onesidedc). Evaluate the following.

1.  $\displaystyle \lim_{x\to 1^-} f(x)$
1.  $\displaystyle \lim_{x\to 1^+} f(x)$
1.  $\displaystyle \lim_{x\to 1} f(x)$
1.  $f(1)$


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/Nn6JoJRK7nk")
```


```{r onesidedc, fig.cap="The function for Example \\@ref(ex:onesidec)", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/rjr9ejfd")
```

It is clear by looking at the graph that both the left and right-hand limits of $f$, as $x$ approaches 1, are 0. Thus it is also clear that _the_ limit is 0; i.e., $\displaystyle \lim_{x\to 1} f(x) = 0$. It is also clearly stated that $f(1) = 1$.

### Example: Evaluating limits of a piecewise--defined function {#ex:onesided}

Let $f(x) = \begin{cases} x^2 & 0\leq x\leq 1 \\ 2-x & 1<x\leq 2\end{cases},$ as shown in Figure \ref{fig:onesidedd}. Evaluate the following. 

1.  $\displaystyle \lim_{x\to 1^-} f(x)$
1.  $\displaystyle \lim_{x\to 1^+} f(x)$
1.  $\displaystyle \lim_{x\to 1} f(x)$
1.  $f(1)$


```{r, out.extra='allow="fullscreen; accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"'}
knitr::include_url("https://www.youtube.com/embed/HVFazve-Qxc")
```



```{r onesidedd, fig.cap="The function for Example \\@ref(ex:onesided)", echo=FALSE, out.extra='allow="fullscreen"'}
knitr::include_url("https://www.geogebra.org/calculator/ryxpsrqc")
```

It is clear from the definition of the function and its graph that all of the following are equal:
$$ \lim_{x\to 1^-} f(x) = \lim_{x\to 1^+} f(x) =\lim_{x\to 1} f(x) =f(1) = 1.$$

-----


In Examples \@ref(ex:onesidea) – \@ref(ex:onesided) we were asked to find
both $\displaystyle \lim_{x\to 1}f(x)$ and $f(1)$. Consider the following table:

  Example  | $\displaystyle \lim_{x\to 1}f(x)$ | $f(1)$
  :-----------------:  | :------------------:  | :------------:
  \@ref(ex:onesidea)   |    does not exist     |       1
  \@ref(ex:onesideb)   |          1            |   not defined
  \@ref(ex:onesidec)   |          0            |       1
  \@ref(ex:onesided)   |          1            |       1

Only in Example \@ref(ex:onesided) do both the function and the limit
exist and agree. This seems "nice;" in fact, it seems "normal." This is
in fact an important situation which we explore in the next section,
entitled "Continuity." In short, a *continuous function* is one in which
when a function approaches a value as $x\rightarrow c$ (i.e., when
$\displaystyle \lim_{x\to c} f(x) = L$), it actually *attains* that value at $c$.
Such functions behave nicely as they are very predictable.
