--- 
title: "APEX Calculus 1"
author: "Gregory Hartman, with additional content by Sean Fitzpatrick, Grady Weyenberg, and others"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This text is Part I  of a three-text series on Calculus. This part covers the material typically taught in a first course on Calculus: Limits, the Derivative, and the Integral."
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
library(tidyverse)

eng_ggb <- function(options){
  `%n%` <- knitr:::`%n%`
  options$echo <- FALSE
  options$results <- 'asis'
  options$fig.num <- 1L
  options$fig.show <- 'asis'
  
  code <- yaml::yaml.load(options$code)
  code$id <- code$id %n% options$label
  code$width <- code$width %n% options$width %n% 770
  code$height <- code$height %n% options$height %n% 400
  code$AppName <- code$AppName %n% "graphing"
  options$fig.cap <- options$fig.cap %n% code$caption 
  
  options$json <- jsonlite::toJSON(code, auto_unbox = TRUE)
  
  out = if (options$eval && knitr:::is_html_output(excludes = 'markdown')){
    knitr:::add_html_caption(options,
    knitr:::one_string(c(
      stringr::str_interp('<div id="${label}"></div>', options),
      '<script type="text/javascript">', 
      stringr::str_interp('window.addEventListener("load", function(){
                          (new GGBApplet(${json}, true)).inject("${label}");
                          });', options),
      '</script>'
    )))
  }
  knitr::engine_output(options, options$code, out)
}
knitr::knit_engines$set(geogebra = eng_ggb)

eng_youtube <- function(options){
  `%n%` <- knitr:::`%n%`
  options$echo <- FALSE
  options$results <- 'asis'
  
  code <- yaml::yaml.load(options$code)
  code$width <- code$width %n% options$width %n% 640
  code$height <- code$height %n% options$height %n% 385
  code$allow <- code$allow %n% "accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"
  
  template <- '<iframe width="${width}" height="${height}" src="https://www.youtube.com/embed/${id}?modestbranding=1&rel=0" frameborder="0" allow="${allow}" allowfullscreen></iframe>'
  
  out <- if(options$eval && knitr:::is_html_output(excludes = 'markdown')) {stringr::str_interp(template, code)}
  knitr::engine_output(options, options$code, out)
}
knitr::knit_engines$set(youtube = eng_youtube)
```

# Preface {-}

![CC-BY-NC](https://mirrors.creativecommons.org/presskit/buttons/88x31/svg/by-nc.svg)

Copyright © 2018 Gregory Hartman

Bookdown translation and Geogebra plots by Grady Weyenberg. Other video content copyright of respective creators.

Licensed to the public under Creative Commons Attribution-Noncommercial
4.0 International Public License


Thank you for reading this short preface. Allow us to share a few key
points about the text so that you may better understand what you will
find beyond this page.

This text is Part I of a three–text series on Calculus. The first part
covers material taught in many “Calc 1” courses: limits, derivatives,
and the basics of integration, found in Chapters 1 through 6.1. The
second text covers material often taught in “Calc 2:” integration and
its applications, along with an introduction to sequences, series and
Taylor Polynomials, found in Chapters 5 through 8. The third text covers
topics common in “Calc 3” or “multivariable calc:” parametric equations,
polar coordinates, vector–valued functions, and functions of more than
one variable, found in Chapters 9 through 14. All three are available
separately for free at `www.apexcalculus.com`. These three texts are
intended to work together and make one cohesive text, *APEX Calculus*,
which can also be downloaded from the website.

Printing the entire text as one volume makes for a large, heavy,
cumbersome book. One can certainly only print the pages they currently
need, but some prefer to have a nice, bound copy of the text. Therefore
this text has been split into these three manageable parts, each of
which can be purchased for about \$15 at
[Amazon.com](https://www.amazon.com/APEX-Calculus-v4-0/dp/1719219591).

A result of this splitting is that sometimes a concept is said to be
explored in a “later section,” though that section does not actually
appear in this particular text. Also, the index makes reference to
topics and page numbers that do not appear in this text. This is done
intentionally to show the reader what topics are available for study.
Downloading the .pdf of *APEX Calculus* will ensure that you have all
the content.\

## For Students: How to Read this Text {-}

Mathematics textbooks have a reputation for being hard to read.
High–level mathematical writing often seeks to say much with few words,
and this style often seeps into texts of lower–level topics. This book
was written with the goal of being easier to read than many other
calculus textbooks, without becoming too verbose.

Each chapter and section starts with an introduction of the coming
material, hopefully setting the stage for “why you should care,” and
ends with a look ahead to see how the just–learned material helps
address future problems.

*Please read the text;* it is written to explain the concepts of
Calculus. There are numerous examples to demonstrate the meaning of
definitions, the truth of theorems, and the application of mathematical
techniques. When you encounter a sentence you don’t understand, read it
again. If it still doesn’t make sense, read on anyway, as sometimes
confusing sentences are explained by later sentences.

*You don’t have to read every equation.* The examples generally show
“all” the steps needed to solve a problem. Sometimes reading through
each step is helpful; sometimes it is confusing. When the steps are
illustrating a new technique, one probably should follow each step
closely to learn the new technique. When the steps are showing the
mathematics needed to find a number to be used later, one can usually
skip ahead and see how that number is being used, instead of getting
bogged down in reading how the number was found.

*Most proofs have been omitted.* In mathematics, *proving* something is
always true is extremely important, and entails much more than testing
to see if it works twice. However, students often are confused by the
details of a proof, or become concerned that they should have been able
to construct this proof on their own. To alleviate this potential
problem, we do not include the proofs to most theorems in the text. The
interested reader is highly encouraged to find proofs online or from
their instructor. In most cases, one is very capable of understanding
what a theorem *means* and *how to apply it* without knowing fully *why*
it is true.\

## Interactive, 3D Graphics  {-}

New to Version 3.0 was the addition of interactive, 3D graphics in the
.pdf version. Nearly all graphs of objects in space can be rotated,
shifted, and zoomed in/out so the reader can better understand the
object illustrated.

As of this writing, the only pdf viewers that support these 3D graphics
are Adobe Reader & Acrobat (and only the versions for PC/Mac/Unix/Linux
computers, not tablets or smartphones). To activate the interactive
mode, click on the image. Once activated, one can click/drag to rotate
the object and use the scroll wheel on a mouse to zoom in/out. (A great
way to investigate an image is to first zoom in on the page of the pdf
viewer so the graphic itself takes up much of the screen, then zoom
inside the graphic itself.) A CTRL-click/drag pans the object left/right
or up/down. By right-clicking on the graph one can access a menu of
other options, such as changing the lighting scheme or perspective. One
can also revert the graph back to its default view. If you wish to
deactivate the interactivity, one can right-click and choose the
“Disable Content” option.\

## Thanks {-}

There are many people who deserve recognition for the important role
they have played in the development of this text. First, I thank
Michelle for her support and encouragement, even as this “project from
work” occupied my time and attention at home. Many thanks to Troy
Siemers, whose most important contributions extend far beyond the
sections he wrote or the 227 figures he coded in Asymptote for 3D
interaction. He provided incredible support, advice and encouragement
for which I am very grateful. My thanks to Brian Heinold and Dimplekumar
Chalishajar for their contributions and to Jennifer Bowen for reading
through so much material and providing great feedback early on. Thanks
to Troy, Lee Dewald, Dan Joseph, Meagan Herald, Bill Lowe, John David,
Vonda Walsh, Geoff Cox, Jessica Libertini and other faculty of VMI who
have given me numerous suggestions and corrections based on their
experience with teaching from the text. (Special thanks to Troy, Lee &
Dan for their patience in teaching Calc III while I was still writing
the Calc III material.) Thanks to Randy Cone for encouraging his tutors
of VMI’s Open Math Lab to read through the text and check the solutions,
and thanks to the tutors for spending their time doing so. A very
special thanks to Kristi Brown and Paul Janiczek who took this
opportunity far above & beyond what I expected, meticulously checking
every solution and carefully reading every example. Their comments have
been extraordinarily helpful. I am also thankful for the support
provided by Wane Schneiter, who as my Dean provided me with extra time
to work on this project. Finally, a huge heap of thanks is to be
bestowed on the numerous people I do not know who took the time to email
me corrections and suggestions. I am blessed to have so many people give
of their time to make this book better.\

## APEX -- Affordable Print and Electronic teXts {-}

APEX is a consortium of authors who collaborate to produce high–quality,
low–cost textbooks. The current textbook–writing paradigm is facing a
potential revolution as desktop publishing and electronic formats
increase in popularity. However, writing a good textbook is no easy
task, as the time requirements alone are substantial. It takes countless
hours of work to produce text, write examples and exercises, edit and
publish. Through collaboration, however, the cost to any individual can
be lessened, allowing us to create texts that we freely distribute
electronically and sell in printed form for an incredibly low cost.
Having said that, nothing is entirely free; someone always bears some
cost. This text “cost” the authors of this book their time, and that was
not enough. *APEX Calculus* would not exist had not the Virginia
Military Institute, through a generous Jackson–Hope grant, given the
lead author significant time away from teaching so he could focus on
this text.

Each text is available as a free .pdf, protected by a Creative Commons
Attribution - Noncommercial 4.0 copyright. That means you can give the
.pdf to anyone you like, print it in any form you like, and even edit
the original content and redistribute it. If you do the latter, you must
clearly reference this work and you cannot sell your edited work for
money.

We encourage others to adapt this work to fit their own needs. One might
add sections that are “missing” or remove sections that your students
won’t need. The source files can be found at `github.com/APEXCalculus`.

You can learn more at `www.vmi.edu/APEX`.\

## Version 4.0 {-}

Key changes from Version 3.0 to 4.0:

-   Numerous typographical and “small” mathematical corrections (again,
    thanks to all my close readers!).

-   “Large” mathematical corrections and adjustments. There were a
    number of places in Version 3.0 where a definition/theorem was not
    correct as stated. See `www.apexcalculus.com` for more information.

-   More useful numbering of Examples, Theorems, etc. “Definition
    11.4.2” refers to the second definition of Chapter 11, Section 4.

-   The addition of Section 13.7: Triple Integration with Cylindrical
    and Spherical Coordinates

-   The addition of Chapter 14: Vector Analysis.



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
